---
title: "dataviz hypothesis"
output: html_document
date: "2026-08-19"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggrepel)
library(geosphere)
```

```{r spc-indicators}
# rename the four columns we use, once, so everything below reads plainly
climate <- read.csv("climate_change.csv") %>%
  rename(country   = Pacific.Island.Countries.and.territories,
         year      = TIME_PERIOD,
         value     = OBS_VALUE,
         indicator = CLIMATE_CHANGE_INDICATORS)

oni <- read.csv("oni.csv") %>% select(year = Year, oni = MEAN)

sea_temp <- filter(climate, indicator == "SST_ANOM")   # degrees C, 1850-2025
rain     <- filter(climate, indicator == "RAIN_ANOM")  # mm, 1979-2025
sea_lvl  <- filter(climate, indicator == "SEA_LVL")    # metres, 1993-2023
stations <- filter(climate, indicator == "METEO_MONITOR_NET")

# one row per country-year, always. this is the check that catches the old bug
stopifnot(nrow(count(sea_temp, country, year) %>% filter(n > 1)) == 0)
```

```{r crosswalk}
# GHCN-D and IGRA key on FIPS-2 (first two characters of the station ID);
# WDQMS uses ISO3. Everything joins through this table.
xwalk <- tribble(
  ~fips, ~iso3,  ~country,                    ~grp,
  "PP",  "PNG",  "Papua New Guinea",          "PICT",
  "BP",  "SLB",  "Solomon Islands",           "PICT",
  "NH",  "VUT",  "Vanuatu",                   "PICT",
  "NC",  "NCL",  "New Caledonia",             "PICT",
  "FJ",  "FJI",  "Fiji",                      "PICT",
  "TN",  "TON",  "Tonga",                     "PICT",
  "WS",  "WSM",  "Samoa",                     "PICT",
  "AQ",  "ASM",  "American Samoa",            "PICT",
  "NE",  "NIU",  "Niue",                      "PICT",
  "CW",  "COK",  "Cook Islands",              "PICT",
  "TV",  "TUV",  "Tuvalu",                    "PICT",
  "KR",  "KIR",  "Kiribati",                  "PICT",
  "NR",  "NRU",  "Nauru",                     "PICT",
  "TL",  "TKL",  "Tokelau",                   "PICT",
  "RM",  "MHL",  "Marshall Islands",          "PICT",
  "FM",  "FSM",  "Micronesia",                "PICT",
  "PS",  "PLW",  "Palau",                     "PICT",
  "GQ",  "GUM",  "Guam",                      "PICT",
  "CQ",  "MNP",  "Northern Mariana Islands",  "PICT",
  "FP",  "PYF",  "French Polynesia",          "PICT",
  "WF",  "WLF",  "Wallis and Futuna",         "PICT",
  "PC",  "PCN",  "Pitcairn",                  "PICT",
  "AS",  "AUS",  "Australia",                 "Comparator",
  "NZ",  "NZL",  "New Zealand",               "Comparator"
)

pict       <- filter(xwalk, grp == "PICT")$country
ACTIVE_CUT <- 2021
```

```{r load-archives}
igra <- read_fwf(
  "data/igra2-station-list.txt",
  fwf_positions(c(1, 13, 22, 32, 42, 73, 78, 83),
                c(11, 20, 30, 37, 71, 76, 81, 88),
                c("id", "lat", "lon", "elev", "name", "first", "last", "nobs")),
  col_types = "cddd" %>% paste0("ciii")
) %>%
  mutate(fips = str_sub(id, 1, 2)) %>%
  inner_join(xwalk, by = "fips")

ghcn_stn <- read_fwf(
  "data/ghcnd-stations.txt",
  fwf_positions(c(1, 13, 22, 32, 42), c(11, 20, 30, 37, 71),
                c("id", "lat", "lon", "elev", "name")),
  col_types = "cdddc"
)

ghcn_prcp <- read_fwf(
  "data/ghcnd-inventory.txt",
  fwf_positions(c(1, 13, 22, 32, 37, 42), c(11, 20, 30, 35, 40, 45),
                c("id", "lat", "lon", "elem", "first", "last")),
  col_types = "cddcii"
) %>%
  filter(elem == "PRCP") %>%
  mutate(fips = str_sub(id, 1, 2)) %>%
  inner_join(xwalk, by = "fips") %>%
  left_join(select(ghcn_stn, id, elev, name), by = "id")

stopifnot(nrow(igra) > 0, nrow(ghcn_prcp) > 0)
```

```{r load-wdqms}
# ---------------------------------------------------------------------
# WDQMS status categories:
#   green   normal / complete
#   purple  more than declared in OSCAR - schedule metadata fault, but
#           the data IS arriving, so counted as reporting
#   orange  availability issue, >= 30% received
#   red     availability issue, < 30% received
#   black   nothing received in the period
#   grey    OSCAR schedule issue      } metadata faults,
#   yellow  no match in OSCAR/Surface }  excluded from every rate
#
# Each station appears once per monitoring centre (DWD, ECMWF, JMA,
# NCEP). WDQMS shows the best result across centres; collapse the same way.
# ---------------------------------------------------------------------
status_rank <- c(green = 1, purple = 2, orange = 3, red = 4,
                 grey = 5, yellow = 6, black = 7)

# Pacific territories report under the administering country's ISO code.
# Guam and American Samoa appear as USA; without this they vanish from
# the per-country tables and appear to have no stations at all.
reassign_territory <- function(df) {
  mutate(df, iso3 = case_when(
    str_detect(name, regex("GUAM", ignore_case = TRUE))                     ~ "GUM",
    str_detect(name, regex("PAGO PAGO|AMERICAN SAMOA", ignore_case = TRUE)) ~ "ASM",
    str_detect(name, regex("SAIPAN|TINIAN|ROTA", ignore_case = TRUE))       ~ "MNP",
    str_detect(name, regex("WALLIS|FUTUNA|HIHIFO", ignore_case = TRUE))     ~ "WLF",
    TRUE ~ iso3))
}

read_wdqms <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    rename(status = `color code`, iso3 = `country code`,
           received = `#received`, expected = `#expected`) %>%
    reassign_territory() %>%
    mutate(rank = status_rank[status]) %>%
    arrange(rank) %>%
    distinct(wigosid, iso3, .keep_all = TRUE) %>%
    inner_join(xwalk, by = "iso3")
}

land  <- read_wdqms("data/wdqms_LandSurface_synop_availability_monthly_pressure_oscar_2026-07.csv")
upper <- read_wdqms("data/wdqms_UpperAir_temp_availability_monthly_oscar_2026-07.csv")

# the marine quality export reports `rms` - root-mean-square departure
# of the observation from the model background, in hPa
marine <- read_csv("data/wdqms_marine_surface_quality_monthly_pressure_2026-07.csv",
                   show_col_types = FALSE) %>%
  rename(departure = rms)

reporting_rate <- function(df, layer) {
  df %>%
    filter(!status %in% c("grey", "yellow")) %>%
    group_by(country, grp) %>%
    summarise(registered = n(),
              reporting  = sum(status %in% c("green", "purple")),
              silent     = sum(status == "black"),
              received   = sum(received),
              expected   = sum(expected), .groups = "drop") %>%
    mutate(pct_obs = if_else(expected > 0, round(100 * received / expected), NA_real_),
           layer   = layer)
}
```

# 1. Start with you

*"When and where were you born?"*

```{r birthplace}
birth_year <- 1990
place      <- "Nauru"

mine <- filter(sea_temp, country == place)
then <- mean(mine$value[mine$year >= birth_year & mine$year < birth_year + 10])
now  <- mean(mine$value[mine$year >= 2016])

round(now - then, 1)
```

```{r birthplace-plot}
ggplot(mine, aes(year, value)) +
  geom_line(colour = "#1a3d5c") +
  labs(title = paste("Since", birth_year, ", seas around ", place, "have warmed"),
       x = NULL, y = "Sea temperature anomaly (C)") +
  theme_minimal()
```

# 2. The long record was never about these islands

Ships logged sea temperature on their way to somewhere else, and
satellites read the surface from orbit. Both cover the Pacific
incidentally. Neither was set up to watch these islands.

That matters for everything below. A line drawn back to 1850 is not the
same thing as a country measuring its own weather, and the two come
apart badly once we look at what is on the ground.

```{r warming-hinge}
# average the countries so the shared signal shows through
yearly <- sea_temp %>% group_by(year) %>% summarise(temp = mean(value))

# a "hinge" is 0 before its year and counts years after it,
# so adding two of them lets the slope change twice
hinge <- function(x, b) pmax(0, x - b)

fit <- lm(temp ~ year + hinge(year, 1915) + hinge(year, 1985), yearly)

# the coefficients are changes in slope, so add them up as you go
round(cumsum(coef(fit)[-1]) * 10, 2)      # degrees C per decade, per era
```

```{r warming-hinge-plot}
yearly$fitted <- predict(fit)

ggplot(yearly, aes(year)) +
  geom_line(aes(y = temp), colour = "grey70") +
  geom_line(aes(y = fitted), colour = "#1a3d5c", linewidth = 1.2) +
  geom_vline(xintercept = c(1915, 1985), linetype = "dotted") +
  labs(title = "The warming tripled in the mid-1980s",
       x = NULL, y = "Sea temperature anomaly (C)") +
  theme_minimal()
```

Sea temperatures are rising alongside sea levels, and every country is
affected.

```{r all-of-them}
warming <- sea_temp %>% filter(year >= 1985) %>% group_by(country) %>%
  summarise(per_decade = coef(lm(value ~ year))[2] * 10)

rising <- sea_lvl %>% group_by(country) %>%
  summarise(mm_per_year = coef(lm(value ~ year))[2] * 1000)

# Deliberately not a chart. The finding is "all of them", and four numbers say
# that more plainly than 22 near-identical upward lines would. Drop the chunk
# entirely if the sentence above carries it.
c(countries_warming = sum(warming$per_decade > 0), of = nrow(warming),
  countries_rising  = sum(rising$mm_per_year > 0), of = nrow(rising))
```

# 3. Rainfall is the signal that hurts, and it swings

Temperature is the thing the long record measures well. Rainfall is the
thing people plan around, and it behaves differently. It does not climb.
It lurches: a year of too much, then a year of nothing.

```{r rain-ratio}
window   <- c(1985, max(rain$year))
rain_win <- filter(rain, year >= window[1])

jump  <- rain_win %>% group_by(country) %>% summarise(s = sd(value))
climb <- rain_win %>% group_by(country) %>%
  summarise(t = coef(lm(value ~ year))[2] * 10)   # mm per decade

# median ABSOLUTE trend - the signed median lets wetting and drying countries
# cancel, which is what inflated this to 8.5
trend <- median(abs(climb$t))
ratio <- median(jump$s) / trend

round(median(jump$s), 1)   # 11.9
round(trend, 1)            # 2.5
round(ratio, 1)            # 4.7
```

From one year to the next, rainfall moves about `r round(ratio, 0)`
times as far as its trend moves in a decade (1985–2025). Year-to-year
movement here is the median country standard deviation
(`r round(median(jump$s), 1)` mm); trend is the median absolute
per-decade trend (`r round(trend, 1)` mm).

```{r rain-lurch-plot}
ggplot(rain |> group_by(year) |> summarise(value=mean(value)),
              aes(year, value)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(colour = "#eb6834") +
  labs(title = paste("Rainfall in the Pacific does not trend - it lurches"),
       x = NULL, y = "Rainfall anomaly (mm)") +
  theme_minimal()
```

The temperature trend was only measurable because the same places were
measured the same way for decades. Rainfall asks for more than that. It
moves faster and over shorter distances, so seeing it takes more
instruments in more places — and it is the thing the region has fewest
instruments for.

The swings are not random, which is what makes them forecastable in
principle. During El Niño, warm water slides east and the rain follows
it, so one half of the Pacific is wet while the other is dry.

```{r spcz-grouping}
# Assignment by geography, fixed before looking at the rainfall series, so the
# El Nino effect is tested against an external map rather than derived from the
# same index that defines it.
#
# MECHANISM: Brown, Lengaigne, Lintner, Widlansky, van der Wiel, Dutheil,
# Linsley, Matthews & Renwick (2020), "South Pacific Convergence Zone dynamics,
# variability and impacts in a changing climate", Nature Reviews Earth &
# Environment. Islands near the equator and east of the mean SPCZ get wetter;
# islands south-west of it get drier.
#
# PER-COUNTRY: BoM/CSIRO PACCSAP country brochures, pacificclimatechangescience.org,
# except New Caledonia (Meteo-France) and Tokelau (NIWA Island Climate Update).
# Every quote below was checked against its source document. Note how many are
# station-specific - Suva, Nuku'alofa, Funafuti, Port Vila - which is the
# station argument turning up inside this section's own sourcing.
spcz_itcz <- c(
  # south-west Pacific - the SPCZ moves away from them
  'Papua New Guinea' = 'dry',   # "El Nino years are usually drier than normal"
  'Solomon Islands'  = 'dry',   # "warmer, drier wet season conditions"
  'New Caledonia'    = 'dry',   # Meteo-France: "un risque accru de secheresse"
  'Vanuatu'          = 'dry',   # Port Vila, Aneityum: "tend to bring drier conditions"
  'Fiji'             = 'dry',   # Suva: dry seasons "drier and cooler than normal"
  'Tonga'            = 'dry',   # Nuku'alofa, Lupepau'u: "drier wet seasons than normal"
  'Samoa'            = 'dry',   # "wet seasons that are drier than normal"
  'Niue'             = 'dry',   # "drier than normal, particularly in the wet season"
  # equatorial central Pacific - the convection moves over them
  'Kiribati'         = 'wet',   # "Across Kiribati... wetter, warmer conditions than normal"
  'Tuvalu'           = 'wet',   # Funafuti: "wetter, warmer conditions than normal"
  'Nauru'            = 'wet',   # "warmer, wetter conditions than normal"
  'Tokelau'          = 'wet'    # NIWA ICU 111 (Dec 2009): "enhanced convection... near Tokelau"
)
# Two corrections made on checking the sources:
#   - the earlier Tuvalu quote ("the SPCZ tends to move to the north-east over
#     Tuvalu") is NOT in the brochure. Direction is right; quote replaced with
#     the Funafuti sentence, which is.
#   - the Kiribati brochure says "Across Kiribati" and does not split Gilberts
#     from Line Islands. That distinction is Brown et al., not the brochure.

wet_side <- names(spcz_itcz)[spcz_itcz == "wet"]
dry_side <- names(spcz_itcz)[spcz_itcz == "dry"]
# Ten countries left out, each for a reason - they have no single El Nino
# rainfall direction, so averaging them in would blur the split rather than
# test it.
#   Palau, Micronesia, Marshall Islands, Guam, N. Mariana Islands
#     Micronesia runs on a lag: near or above normal early in the El Nino year,
#     drought Jan-Mar of the FOLLOWING year. An annual mean cancels the two.
#   Cook Islands       Rarotonga drier, northern atolls wetter.
#   French Polynesia   Marquesas wetter, Australes drier, sign flips with flavour.
#   Wallis and Futuna  Meteo-France: ENSO has little significant effect there.
#   American Samoa     Sits on the SPCZ axis; individual events go both ways.
#   Pitcairn           No authoritative ENSO rainfall assessment exists.
excluded <- setdiff(unique(rain$country), names(spcz_itcz))
excluded

lean <- rain %>% inner_join(oni, by = "year") %>% group_by(country) %>%
  summarise(r = cor(value, oni))

check <- lean %>% filter(country %in% names(spcz_itcz)) %>%
  mutate(assigned = spcz_itcz[country],
         agrees   = (assigned == "wet" & r > 0) | (assigned == "dry" & r < 0))

check
mean(check$agrees)   # proportion of data agrees with the map

wet <- rain %>% filter(country %in% wet_side) %>% group_by(year) %>%
  summarise(m = mean(value))   # mean rainfall anomaly, wet-side countries
dry <- rain %>% filter(country %in% dry_side) %>% group_by(year) %>%
  summarise(m = mean(value))   # mean rainfall anomaly, dry-side countries

# Check before writing prose around the chart. Brown et al. note that in the
# strongest eastern-Pacific events - 1982-83 and 1997-98 - Nauru and Tarawa
# went DRY despite normally being wetter, because the convection overshoots
# the Gilberts and sits over the Line Islands.
rain %>% filter(country %in% wet_side, year %in% c(2015, 2016)) %>%
  select(country, year, value) %>% arrange(country, year)
```

```{r split-gap}
split <- inner_join(wet, dry, by = "year", suffix = c("_wet", "_dry")) %>%
  mutate(gap = m_wet - m_dry)

filter(split, year %in% c(2015, 2016))   # the event this piece follows
round(median(abs(split$gap)), 1)         # a normal year, for scale
```

```{r split-plot}
# Only 2015-16 is shaded. An earlier version also shaded 1999-2000 and labelled
# it an El Nino - that was a strong La Nina, so it has been removed rather than
# corrected, since the piece follows one event.
sides <- bind_rows(mutate(wet, side = "wetter in El Nino"),
                   mutate(dry, side = "drier in El Nino"))

ggplot(sides, aes(year, m, colour = side)) +
  annotate("rect", xmin = 2014.5, xmax = 2016.5,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = .55) +
  annotate("text", x = 2015.5, y = Inf, label = "2015-16",
           vjust = 1.6, size = 3, colour = "grey30") +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("wetter in El Nino" = "#2a78d6",
                                 "drier in El Nino"  = "#d03b3b")) +
  labs(title = "In an El Nino, the Pacific splits in half",
       subtitle = "one side floods while the other dries out",
       x = NULL, y = "Rainfall anomaly (mm)", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

So the mechanism is known and the pattern holds. But a farmer does not
plant against a pattern for half an ocean. What has to be forecast is
how much rain falls in one district, and when it stops — and that is a
question you can only answer from the record of that district.

# 4. What that cost: Papua New Guinea, 2015

In Papua New Guinea the rain began falling away in April 2015. Between
mid-July and mid-August, at the one high-altitude research station
recording it, temperatures fell to minus two degrees before dawn. The
frosts destroyed sweet potato, the staple, in gardens above 2,200 metres
— where a replanted crop takes eight to twelve months to mature.[^pb11-frost]

[^pb11-frost]: R.M. Bourke, Bryant Allen and Michael Lowe, *Estimated
    impact of drought and frost on food supply in rural PNG in 2015*,
    Development Policy Centre Policy Brief 11, Australian National
    University, January 2016, pp. 1–2. "Since April 2015, much of rural
    Papua New Guinea (PNG) has been severely impacted by a severe
    drought and, at a number of very high altitude locations above
    2200 m altitude, by repeated frosts." Frost temperatures of −2 °C
    were recorded between 5 and 7 am at Tambul High Altitude
    Agricultural Research Station between mid-July and mid-August 2015
    (Kud Sitango, NARI Tambul, pers. comm.). Sweet potato takes 8–12
    months to mature above 2,200 m (p. 5).

By the time anyone tried to count, food was very scarce or extremely
scarce in 27 of Papua New Guinea's 271 rural districts. Those districts
hold about 770,000 people, though the assessment calls that an upper
limit and says the number actually short of food is likely to be
considerably less, since not everyone in a district was equally
affected. Nearly nine in ten of them live in the highlands.[^pb11-assess]

[^pb11-assess]: Bourke, Allen and Lowe, op. cit., pp. 2, 4–5 and
    Tables 2–3. 27 of 271 rural LLGAs at rating 4 or 5; a maximum of
    770,000 people, which the authors describe as "an upper limit" with
    the number very short of food "likely to be considerably less"; 53%
    at very high altitude and 35% in the central highlands or highlands
    fringe; food supply information obtained for 141 LLG areas and
    extrapolated to the remaining 130; "the only national assessment of
    the food supply situation for all PNG at the LLGA level".

That assessment was built from more than two hundred reports — formal
surveys, emails, phone calls, newspaper stories, photographs posted on
Facebook. It covered 141 of the 271 districts. The other 130 were
estimated from what had happened nearby, and from what had happened in
1997. It remains the only national account of the emergency at that
resolution. Its authors asked for the numbers to be checked in the
field. That was not done.[^dfat-check]

[^dfat-check]: DFAT Office of Development Effectiveness, *Evaluation of
    Australia's response to the PNG El Niño drought and frosts 2015–17*:
    "The researchers recommended field checking to clarify actual
    numbers and set priorities; but this was not undertaken." Verified
    against the document.

There is no official death toll. Australia's review of its own response
says an increase in mortality is likely and impossible to
quantify.[^dfat-mortality]

[^dfat-mortality]: DFAT evaluation, op. cit.: "It is likely that there
    was some increase is [sic] all-causes mortality due to the 2015 El
    Niño… But such an increase is impossible to quantify." Verified
    against the document.

```{r png-coverage-plot}
# Coverage. The brief obtained food supply information for 141 of the 271
# rural districts and extrapolated the other 130 from neighbouring areas and
# from what happened in 1997. This is the chart the section rests on.
coverage <- data.frame(i = 0:270) %>%
  mutate(row = i %/% 30, col = i %% 30,
         status = ifelse(i < 141, "Food supply reported",
                                  "Estimated from neighbouring areas and from 1997"))

ggplot(coverage, aes(col, -row, fill = status)) +
  geom_tile(colour = "#1a3d5c", linewidth = .35, width = .85, height = .85) +
  scale_fill_manual(values = c("Food supply reported" = "#1a3d5c",
    "Estimated from neighbouring areas and from 1997" = "white")) +
  coord_equal() +
  labs(title = "Papua New Guinea's 271 rural districts, 2015 drought assessment",
       fill = NULL) +
  theme_void() + theme(legend.position = "bottom")
```

```{r png-llga}
# Table 2 of Policy Brief 11, transcribed by hand from the PDF.
# CHECK IT: these columns sum to 752,234 against the brief's published
# subtotal of 752,227, so at least one figure is misread.
llg <- read.csv("data/png_2015_cat45_llga.csv")
```

```{r png-zones-plot}
# Where the affected population was. Milne Bay is hard-coded because it is
# Table 3 of the brief, which is not in the csv - add those seven island
# wards if this chart is used.
zones <- llg %>% group_by(zone) %>% summarise(pop = sum(pop_2015_est)) %>%
  bind_rows(data.frame(zone = "Small islands, Milne Bay", pop = 18705)) %>%
  mutate(share = pop / sum(pop),
         highlands = zone %in% c("Very high altitude", "Highland and highland fringe"))

ggplot(zones, aes(pop / 1000, reorder(zone, pop), fill = highlands)) +
  geom_col(width = .65) +
  geom_text(aes(label = sprintf("%.0fk  (%.0f%%)", pop / 1000, share * 100)),
            hjust = -.15, size = 3.1, colour = "grey35") +
  scale_fill_manual(values = c("TRUE" = "#1a3d5c", "FALSE" = "grey78"), guide = "none") +
  expand_limits(x = 500) +
  labs(title = "Nearly nine in ten were in the highlands",
       x = "Thousand people in category 4 or 5 districts", y = NULL) +
  theme_minimal()
```

```{r png-districts-plot}
# The 27 districts themselves.
ggplot(llg, aes(pop_2015_est / 1000, reorder(llga, pop_2015_est),
                colour = zone, size = factor(rating))) +
  geom_point() +
  scale_size_manual(values = c("4" = 2.2, "5" = 4), name = "Rating") +
  scale_colour_manual(values = c("Very high altitude" = "#1a3d5c",
                                 "Highland and highland fringe" = "#5b8ab0",
                                 "Inland lowland Western Province" = "#eb6834"),
                      name = NULL) +
  labs(title = "The 27 districts where food was very or extremely scarce",
       subtitle = "rating 5 = no food available at all; 4 = famine food only",
       x = "Population, 2015 estimate (thousands)", y = NULL) +
  theme_minimal() + theme(legend.position = "bottom", legend.box = "vertical")
```

The ocean was watched. By early April 2015, forecasters including the
Australian Bureau of Meteorology were reporting a severe El Niño moving
west.[^dfat-april] What no one could say was how much rain would fall in
the Kandep basin, or when it would stop.

[^dfat-april]: DFAT evaluation, op. cit.: "By early April 2015 climate
    scientists, including from the Australian Bureau of Meteorology,
    were reporting on the likelihood of a severe El Niño event, moving
    from the eastern Pacific westwards." Verified against the document.

An El Niño declaration is a statement about the ocean. A rainfall
forecast for a district is a different thing: it is built from what
rainfall at that place has done in past El Niños, so it can only ever be
as local as the record behind it. The highlands are where most of Papua
New Guinea's people live, between roughly 1,500 and 2,800 metres. They
are also where almost none of the measurement is.

# 5. Rainfall takes three layers

Rainfall is not one measurement. It takes three, and they do different
jobs.

- **The ocean surface** — buoys and ships — shows where the warm water
  and the moisture sit, weeks to seasons ahead.
- **The upper air** — radiosondes on balloons — shows where the
  convection will go over the next few days.
- **The ground** — rain gauges — records what actually arrived, and
  calibrates everything else, including the satellites.

In the Pacific, one of these works, one is thin and misplaced, and one
has effectively gone.

## The ocean layer works

```{r marine-quality}
pacific_box <- marine %>%
  filter(between(latitude, -30, 25), longitude >= 130 | longitude <= -130)

tibble(platforms_pacific   = nrow(pacific_box),
       platforms_global    = nrow(marine),
       median_rms_pacific  = round(median(pacific_box$departure), 2),
       median_rms_global   = round(median(marine$departure), 2),
       pct_within_1hPa_pac = round(100 * mean(pacific_box$departure <= 1)),
       pct_within_1hPa_glb = round(100 * mean(marine$departure <= 1)))

count(pacific_box, marine_type)
```

`r nrow(pacific_box)` platforms reported in the tropical Pacific through
July 2026 — moorings along the equator, ships on the lanes, drifters
between. `r round(100 * mean(pacific_box$departure <= 1))`% sat within 1
hPa RMS of the forecast model, against
`r round(100 * mean(marine$departure <= 1))`% globally. The ocean around
these islands is better observed than the ocean on average.

Almost none of it is paid for or operated from within the region, which
matters for how long it lasts. It does not matter for whether the data
exists today. It exists.

## The ground layer is at sea level, and half of it is silent

```{r png-elevation}
png_gauges <- filter(ghcn_prcp, country == "Papua New Guinea")

png_summary <- png_gauges %>%
  summarise(ever         = n(),
            active       = sum(last >= ACTIVE_CUT),
            highest_ever = max(elev, na.rm = TRUE),
            above_1000   = sum(elev > 1000, na.rm = TRUE))

png_summary
```

Of Papua New Guinea's `r png_summary$ever` internationally exchanged
rain gauges, `r png_summary$active` are still reporting. The highest
sits at `r round(png_summary$highest_ever)` metres. Across the whole
record, `r png_summary$above_1000` have ever been placed above 1,000
metres — in a country whose population lives, and whose crops failed, at
between 1,500 and 2,800.

```{r png-gauge-elevation-plot, fig.height=3.4}
ggplot(png_gauges, aes(elev, fct_reorder(name, elev))) +
  annotate("rect", xmin = 1500, xmax = 2800, ymin = -Inf, ymax = Inf,
           fill = "#d03b3b", alpha = .12) +
  annotate("text", x = 2150, y = 1.2, label = "where the 2015 drought bit",
           size = 3, colour = "#d03b3b") +
  geom_segment(aes(x = 0, xend = elev, yend = name),
               colour = "grey75", linewidth = .4) +
  geom_point(aes(colour = last >= ACTIVE_CUT), size = 2.4) +
  scale_x_continuous(limits = c(0, 3000)) +
  scale_y_discrete() +
  scale_colour_manual(values = c("TRUE" = "#1a3d5c", "FALSE" = "grey70"),
                      labels = c("TRUE" = "still reporting", "FALSE" = "stopped"),
                      name = NULL) +
  labs(title = "Every rain gauge in Papua New Guinea is at sea level",
       subtitle = "internationally exchanged gauges, GHCN-Daily",
       x = "Elevation (m)", y = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", panel.grid.major.y = element_blank())
```

The highland stations do exist. They are simply not switched on.

```{r png-wdqms}
png_land <- filter(land, country == "Papua New Guinea") %>%
  select(name, received, expected, status) %>%
  arrange(desc(received), name)

png_land
```

Papua New Guinea has `r nrow(png_land)` surface stations registered with
WMO. Across July 2026 they returned
`r format(sum(png_land$received), big.mark = ",")` of
`r format(sum(png_land$expected), big.mark = ",")` expected observations
— `r round(100 * sum(png_land$received) / sum(png_land$expected))`% —
and `r sum(png_land$status == "black")` of them sent nothing at all. Two
of the silent ones are **Goroka** and **Mendi North**: Eastern and
Southern Highlands, at altitude, in the provinces the drought hit
hardest. Goroka returned 0 of 403 expected observations that month.
Mendi North, 0 of 248.

```{=html}
<!--
CAVEAT TO PRINT: GHCN-D holds stations whose data is exchanged
internationally. PNG's national service may hold highland records that
never leave the country. That does not weaken the argument - the regional
forecast centres, the global models and the satellite rainfall
calibration all draw on the internationally exchanged set, so a record
nobody outside PNG can see does not inform the outlook PNG gets back. Say
this before a reviewer does.
-->
```

## The upper-air layer has gone

```{r upper-air-history}
upper_hist <- igra %>%
  group_by(country) %>%
  summarise(sites_ever = n(), active = sum(last >= 2024),
            last_year = max(last), .groups = "drop")

lost  <- filter(upper_hist, country %in% pict, active == 0, sites_ever > 0)
never <- setdiff(pict, igra$country)

arrange(lost, desc(last_year)) %>% select(country, sites_ever, last_year)
never
```

Papua New Guinea's last sounding was in
**`r filter(lost, country == "Papua New Guinea")$last_year`**. Not the
year before the drought, or the year after. The year of it. Solomon
Islands stopped in
`r filter(lost, country == "Solomon Islands")$last_year`, Nauru in
`r filter(lost, country == "Nauru")$last_year`, Tonga in
`r filter(lost, country == "Tonga")$last_year`, Samoa in
`r filter(lost, country == "Samoa")$last_year`.
`r knitr::combine_words(never)` never had one.

```{r upper-air-decline, fig.height=3.6}
pic_sondes <- filter(igra, country %in% pict)

active_by_year <- tibble(year = 1950:2026) %>%
  mutate(sites = map_int(year, ~ sum(pic_sondes$first <= .x & pic_sondes$last >= .x)))

peak      <- slice_max(active_by_year, sites, n = 1, with_ties = FALSE)
now_sites <- filter(active_by_year, year == 2026)$sites

ggplot(active_by_year, aes(year, sites)) +
  geom_area(fill = "#1a3d5c", alpha = .15) +
  geom_line(colour = "#1a3d5c", linewidth = 1) +
  geom_vline(xintercept = 2015, linetype = "dotted", colour = "#d03b3b") +
  annotate("text", x = 2013, y = peak$sites * .85,
           label = "PNG's last sounding, 2015",
           hjust = 1, size = 3, colour = "#d03b3b") +
  labs(title = paste0("Pacific radiosonde sites peaked at ", peak$sites,
                      " in ", peak$year, ", and stand at ", now_sites),
       subtitle = "sites with soundings in each year, IGRA v2",
       x = NULL, y = "Radiosonde sites") +
  theme_minimal()
```

This is the layer that dies first, and it dies quietly. Every launch
consumes a balloon, gas and a sonde. Upper-air observation is not a
thing you install; it is a thing you pay for twice a day, forever, with
trained staff. So it is the first line item to go when a programme ends,
and its ending produces no visible event. The forecast kept arriving —
interpolated from whatever instrument was nearest and still working,
still about right for the region and much less useful for any one valley
— and a record that ran for decades stopped without anyone signing off
on it.

# 6. Existing is not reporting, and who pays decides which

A station in the register is not a station sending data. WMO monitors
both, and the difference between them is where the network quietly
stopped.

```{r reporting-both-layers}
rates <- bind_rows(reporting_rate(land,  "Surface"),
                   reporting_rate(upper, "Upper air"))

rates %>%
  select(country, layer, registered, reporting, silent, received, expected, pct_obs) %>%
  arrange(layer, pct_obs)
```

Papua New Guinea has
`r filter(rates, country == "Papua New Guinea", layer == "Upper air")$registered`
radiosonde stations registered with WMO and received
**`r filter(rates, country == "Papua New Guinea", layer == "Upper air")$received`
of
`r filter(rates, country == "Papua New Guinea", layer == "Upper air")$expected`**
expected soundings in July 2026. Solomon Islands and Tuvalu are also
registered for upper-air observation, and also sent nothing. The
obligation is on the books. The balloons are not going up.

The two layers also come apart *inside the same country*, and that is
the tell.

```{r layer-divergence, fig.height=4.4}
paired <- rates %>%
  filter(!is.na(pct_obs)) %>%
  select(country, grp, layer, pct_obs) %>%
  pivot_wider(names_from = layer, values_from = pct_obs) %>%
  filter(!is.na(Surface), !is.na(`Upper air`))

ggplot(paired, aes(Surface, `Upper air`)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey80", linetype = "dashed") +
  annotate("text", x = 95, y = 88, label = "both layers equal",
           angle = 45, size = 3, colour = "grey55") +
  geom_point(aes(colour = grp), size = 3, alpha = .8) +
  geom_text_repel(aes(label = country), size = 2.7, colour = "grey35",
                  seed = 1, max.overlaps = 20) +
  scale_colour_manual(values = c(PICT = "#1a3d5c", Comparator = "#eb6834"),
                      name = NULL) +
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  labs(title = "The layer somebody else pays for is the layer that works",
       subtitle = "observations received against expected, July 2026",
       x = "Surface stations", y = "Radiosonde stations") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Micronesia returned
`r filter(rates, country == "Micronesia", layer == "Upper air")$pct_obs`%
of its expected soundings and
`r filter(rates, country == "Micronesia", layer == "Surface")$pct_obs`%
of its expected surface observations. Palau,
`r filter(rates, country == "Palau", layer == "Upper air")$pct_obs`%
against
`r filter(rates, country == "Palau", layer == "Surface")$pct_obs`%. The
Marshall Islands,
`r filter(rates, country == "Marshall Islands", layer == "Upper air")$pct_obs`%
against
`r filter(rates, country == "Marshall Islands", layer == "Surface")$pct_obs`%.

These are the same countries, the same staff, the same islands. What
differs is that the radiosonde programme in the freely associated states
is run and funded through the United States National Weather Service,
while the surface network is a national responsibility. Where an outside
programme carries the recurring cost, the observations arrive. Where it
does not, they do not — and the pattern reverses cleanly for Papua New
Guinea, which is on its own for both.

That is not a comment on anyone's competence. It is a statement about
which line items have a funder, and it says where money would actually
change something.

It is also what makes the gap so easy to miss. Every silent station is
still counted. They appear in the register, in the network totals, and in
any assessment built on how many stations a country has. On paper the
region is better observed than it was thirty years ago.

# 7. Close

The same administrative logic decides what *is* watched. Sea level is
measured carefully in places with no working weather station at all —
not because anyone judged one threat to matter more, but because tide
gauges and rain gauges are different instruments, installed by different
programmes, on different funding cycles.

```{r sea-level-vs-weather}
latest  <- stations %>% filter(year == max(year)) %>% select(country, n = value)
watched <- latest %>%
  mutate(sea_level = country %in% unique(sea_lvl$country))

sum(watched$sea_level)
filter(watched, n == 0, sea_level)
```

That is worth saying plainly, because it makes this a fixable
administrative fact rather than a natural feature of small islands.

The Pacific can describe its rainfall after the fact, thinly and in the
wrong places. It cannot see it coming. The layer of the atmosphere that
would show it arriving has been switched off across most of the region,
one station at a time, and much of what remains on the ground is
registered, counted, and silent.

*You cannot get ready for a change you cannot watch arriving. How ready
are we for the next El Niño?*

------------------------------------------------------------------------

# Methods and caveats

Two objections a reviewer will raise, answered here rather than in the
narrative.

## "Doesn't the region meet the international standard?"

Mostly, yes — and that is the problem with the standard as a measure of
rainfall forecasting. GBON asks for surface land stations every 200 km,
relaxed to 500 km for small island states, so global forecast models
have enough of the atmosphere sampled to work with. For upper-air
stations the requirement is 500 km, relaxed to 1,000 km for small island
states, with two soundings a day.[^gbon]

[^gbon]: WMO, *GBON Baseline 2023*, INF 6.2 to the Sixth SOFF Steering
    Committee, 27 November 2023, un-soff.org. Maximum spacing under the
    Global Basic Observing Network is given as 200 km for surface land
    stations and 500 km for upper-air stations operated from land,
    relaxed to 500 km and 1,000 km respectively for Small Island States.
    Surface stations are required to share data once an hour and
    upper-air stations to make two soundings a day; a station counts as
    compliant when it reported 80% of required reports on at least 80%
    of days in the assessment month. Note that WMO's GBON FAQ quotes
    Technical Regulation 3.2.2.8 as "horizontal resolutions of 100 km or
    higher" for surface land networks — that is the design aspiration in
    the Technical Regulations, while the figures above are the baseline
    used for gap analysis and for SOFF funding decisions, which is the
    comparison being made here.

```{r gbon-ground}
# 500 km spacing = one per 250,000 km2 =  4 per million km2 (SIDS)
# 200 km spacing = one per  40,000 km2 = 25 per million km2 (everyone else)
# Applied to LAND, which is what the surface land provision covers.
ground <- ghcn_prcp %>%
  group_by(country, grp) %>%
  summarise(gauges_ever = n(),
            active      = sum(last >= ACTIVE_CUT),
            last_obs    = max(last), .groups = "drop") %>%
  mutate(required_per_million = if_else(grp == "PICT", 4, 25))

arrange(ground, desc(active))
```

Run it honestly and the atoll states pass comfortably. Tuvalu is 26
square kilometres; one gauge satisfies 500 km spacing several thousand
times over. Nauru is 21. The rule returns *compliant* and means almost
nothing.

The spacing was chosen for the pressure field, which is smooth and
continental in scale. Rainfall is neither. A network built to keep a
global model stable is doing something different from a network built to
see a drought forming in a valley, and passing the first test says
nothing about the second.

## "Isn't this just density — small countries have few stations?"

Land area is the wrong denominator, because a country is not a tile. It
is a set of places, and the distance between them is what a rain gauge
cannot cross.

```{r dispersion}
max_separation_km <- function(lat, lon) {
  if (length(lat) < 2) return(NA_real_)
  max(distm(cbind(lon, lat)) / 1000)
}

spread <- ghcn_prcp %>%
  filter(last >= ACTIVE_CUT, country %in% pict) %>%
  group_by(country) %>%
  summarise(active = n(), span_km = max_separation_km(lat, lon)) %>%
  arrange(desc(span_km))

spread
```

Kiribati's two reporting gauges sit
`r format(round(filter(spread, country == "Kiribati")$span_km, -1), big.mark = ",")`
km apart. Micronesia's twenty span
`r format(round(filter(spread, country == "Micronesia")$span_km, -1), big.mark = ",")`
km. Papua New Guinea's five are strung along
`r format(round(filter(spread, country == "Papua New Guinea")$span_km, -1), big.mark = ",")`
km of coast with nothing inland.

```{r dispersion-plot, fig.height=4.2}
ggplot(spread, aes(span_km, active)) +
  geom_point(aes(size = active), colour = "#1a3d5c", alpha = .7) +
  geom_text_repel(aes(label = country), size = 2.6, colour = "grey35",
                  seed = 1, max.overlaps = 15) +
  scale_x_continuous(labels = scales::comma) +
  scale_size_continuous(range = c(2, 9), guide = "none") +
  labs(title = "Few gauges, spread across a great deal of ocean",
       subtitle = "gauges still reporting rainfall, GHCN-Daily",
       x = "Greatest distance between two gauges (km)",
       y = "Gauges reporting") +
  theme_minimal()
```

```{r no-gauges}
filter(ground, active == 0, grp == "PICT") %>%
  select(country, gauges_ever, last_obs) %>%
  arrange(last_obs)
```

Every island without one is a place where the satellite rainfall
estimate goes uncalibrated and no forecast can be checked against what
fell.

## Station-count registers disagree

```{r station-counts}
# Settles a claim an earlier draft made and this one does not: that PNG has
# more weather stations than any other country in the region. Three registers
# give three different answers, and PNG tops none of them.
wmo_ind <- stations %>% filter(year == max(year)) %>%
  select(country, wmo_indicator = value)

ghcn_counts <- ghcn_prcp %>% group_by(country) %>%
  summarise(ghcn_ever = n(), ghcn_active = sum(last >= ACTIVE_CUT), .groups = "drop")

# the climate file and the crosswalk name countries differently
# ("Micronesia, Federated State of" vs "Micronesia"), so check before joining
setdiff(wmo_ind$country, ghcn_counts$country)

counts <- full_join(wmo_ind, ghcn_counts, by = "country")

counts %>% slice_max(wmo_indicator, n = 5)
counts %>% slice_max(ghcn_active,   n = 5)
```

```{=html}
<!--
Also for the methods appendix, not yet done: minimum raingauge densities by
physiographic unit - small islands and mountainous terrain are the densest
categories - are in the Guide to Hydrological Practices, WMO-No. 168. Verify
against the current edition before quoting values.

Comparator artefact to state out loud: Australia shares thousands of gauges
with GHCN-D, New Zealand shares around a dozen while operating hundreds. That
is data-sharing policy, not capacity. GHCN-D measures INTERNATIONALLY VISIBLE
rainfall data, which is the right metric here - but only if the piece says so.
-->
```

```{=html}
<!--
OPEN ITEMS

1. The png_2015_cat45_llga.csv transcription is 7 people out against the
   brief's own subtotal (752,234 vs 752,227) - find the misread digit in the
   very-high-altitude or highland-fringe block.
2. Milne Bay (Table 3 of the brief, 7 island wards, 18,705 people) is
   hard-coded in the png-zones-plot chunk rather than in the csv.
3. From the coworker: whether the OCHA situation reports separate frost damage
   from drought. If they conflate the two, that is worth a sentence, because
   section 4 rests on a frost event inside a wider drought.

RESOLVED, for the record:
  - "PNG's National Weather Service declared a severe El Nino on 7 August
    2015" - CUT. That date does not appear in the DFAT evaluation, which gives
    only 22 December 2015, when the National Weather Office declared the PEAK
    had been reached. A first-declaration date needs a PNG NWS source.
  - "why PNG 2015-16 and not another example" - the alternatives (Marshall
    Islands 2016, Palau 2016) are countries excluded from the section 3
    grouping, and the argument here is about localising a forecast rather than
    about who was warned.
  - An earlier draft claimed PNG has the most weather stations in the region.
    It does not; Fiji leads on the WMO indicator. See station-counts above.
-->
```
